\Large Taking uncertainty seriously \newline \normalsize A Bayesian approach to word embedding bias estimation

### 2022-03-15 presentations/presentationBoston/presentationBoston.Rmd
small corrections, added examples

---
title: '\Large Taking uncertainty seriously  \newline \normalsize A  Bayesian approach
  to  word embedding bias estimation '
author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
  group, University of  Gdansk)
date: "Boston, April Fools' Day"
output:
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
  slidy_presentation: default
fontsize: 10pt
classoption: x11names, dvipsnames, bibspacing,natbib, table
urlcolor: blue
bibliography: ../../references/cosineReferences1.bib
csl: ../../references/apa-6th-edition.csl
link-citations: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(ggforce)
library(latex2exp)
th <- theme_tufte(base_size = 20)
options(kableExtra.latex.load_packages = FALSE)
```

@@

\begin{block}{Vectors}
\begin{align*}
a  & = [1,2]\\
b  &= [3,2]
\end{align*}

@@

\begin{align*}
a \cdot b & = a_1 b_1 + a_2 b_2\\
a \cdot a & = a_1^2 + a_2 ^ 2 \\
\lVert a\rVert & = \sqrt(a \cdot a)
\end{align*}

@@

\begin{align*}
a  & = [1,2]\\
b  &= [4,4]
\end{align*}

@@

-  Geometric interpretation: direction (not length)

@@

@@

- ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).

@@

ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT is always maximal in one direction.

@@

@@

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

@@

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t8Plot <- ggplot(t8)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t9Plot <- ggplot(t9)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

>>>>

---
title: '\Large Taking uncertainty seriously  \newline \normalsize A  Bayesian approach
  to  word embedding bias estimation '
author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
  group, University of  Gdansk, Vrij Universiteit Amsterdam)
date: "Boston, April Fools' Day"
output:
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
  slidy_presentation: default
fontsize: 10pt
classoption: x11names, dvipsnames, bibspacing,natbib, table
urlcolor: blue
bibliography: ../../references/cosineReferences1.bib
csl: ../../references/apa-6th-edition.csl
link-citations: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(ggforce)
library(latex2exp)
library(forcats)
th <- theme_tufte(base_size = 20)
options(kableExtra.latex.load_packages = FALSE)
```

@@

\begin{block}{Vectors}
\begin{align*}
a  & = [1,2]\\
b  &= [3,2.5]
\end{align*}

@@

\begin{align*}
a \cdot b & = a_1 b_1 + a_2 b_2\\
a \cdot a & = a_1^2 + a_2 ^ 2 \\
\lVert a\rVert & = \sqrt{(a \cdot a)}
\end{align*}

@@

\begin{align*}
a  & = [1,2]\\
b  &= [3,2.5]
\end{align*}

@@

-  Geometric interpretation: direction $\mathsf{cos} \in (-1,1)$

    - $1$: maximally smilar 
    - $-1$: opposites 
    - $0$: dissimilar

@@

```{r philosophy,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "80%",   message = FALSE, warning = FALSE, results = FALSE}
words <- c("logic", "Nietzsche", "analytic", "Hegel", "Russell", "continental", "burger", "sandwich")
distances <- c(.47, .32, .13, .29, .08, .08, .04, .04)
philosophy <- data.frame(words, distances)
ggplot(philosophy, aes(x = fct_reorder(words, distances), y = distances))+
  geom_point(size = 3)+coord_flip()+theme_tufte(base_size = 20)+ggtitle("Similarity to <philosophy>, examples")+ylab("")+xlab("similarity")
```

### The only "jobs" in top-tens

- Man: robber (.55)

- Woman: policewoman (.6)

@@

- @Ethayarajh2020measuring criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).

@@

- @Ethayarajh2019understanding show that if there are two target words only WEAT is always maximal in one direction.

@@

### Our neutral words (examples, full list size = ADD)

liquor, pow, ballpark, glitchy, billy, dallas, rip, called, outlooks, viet, floater, rattlesnake, exports,
peruvian, recursion, shortfall, corrected, amicable, solutions, diagnostic, patently, flops, approx, percents, lox, catapults,
hamburger, engulfed, households, north, snubbed, playtest

### Our human-related words (examples, full list size = ADD)

switch, studio, stick, soup, sometimes, signal, prior, plant, photo, path, park, near, menu, latter, grass, clock,  wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character, candidate, breakfast, supper, dinner,
eat, drink, carry, run, cast, ask, awake, ear, nose, lunch

@@

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

@@

t1Plot <- ggplot(t1)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B), col = "orangered", width  = 0, height = .02,  size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B),  width  = 0, height = .02, col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t8Plot <- ggplot(t8)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B), width  = 0, height = .02,  col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t9Plot <- ggplot(t9)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B), width  = 0, height = .02,  col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

### 2022-03-15 presentations/presentationBoston/presentationBoston.tex
small corrections, added examples

  pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},

@@

\title{\Large Taking uncertainty seriously \newline \normalsize A
Bayesian approach to word embedding bias estimation}
\author{Alicja Dobrzeniecka \& Rafal Urbaniak
\footnotesize \newline (LoPSE research group, University of Gdansk)}
\date{Boston, April Fools' Day}

@@

\begin{itemize}
\item
  Geometric interpretation: direction (not length)
\item
  \(\mathsf{cosineDistance}\in (0, 2)\)
\item
  Naive interpretation: proximity corresponds to semantic similarity
\end{itemize}
\end{frame}

@@

\begin{block}{No design considerations to sample size}
\protect\hypertarget{no-design-considerations-to-sample-size}{}
\begin{itemize}
\tightlist
\item
  ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues
  that we would need a bias specific dataset of size at least 11903 to
  claim that the system is biased (three times larger than WinoBias).
\end{itemize}

@@

\begin{block}{The form of the definition is suspicious}
\protect\hypertarget{the-form-of-the-definition-is-suspicious}{}
ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT
is always maximal in one direction.

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-2}{}
\begin{block}{Outliers and surprisingly dissimilar words}
\protect\hypertarget{outliers-and-surprisingly-dissimilar-words}{}
We study those by visualizations and uncertainty estimates.
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-3}{}
\begin{block}{Distances for ``muslim''}
\protect\hypertarget{distances-for-muslim}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-4}{}
\begin{block}{Distances for ``priest''}
\protect\hypertarget{distances-for-priest}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-5}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation}{}
\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC (distance) \\
\midrule
\endhead
Biased & 0.859 \\
Hard Debiased & 0.934 \\
Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
\bottomrule
\end{longtable}

@@

>>>>

  pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk, Vrij Universiteit Amsterdam)},

@@

\title{\Large Taking uncertainty seriously \newline \normalsize A
Bayesian approach to word embedding bias estimation}
\author{Alicja Dobrzeniecka \& Rafal Urbaniak
\footnotesize \newline (LoPSE research group, University of Gdansk, Vrij
Universiteit Amsterdam)}
\date{Boston, April Fools' Day}

@@

\begin{itemize}
\item
  Naive interpretation: proximity corresponds to semantic similarity
\item
  Geometric interpretation: direction \(\mathsf{cos} \in (-1,1)\)

  \begin{itemize}
  \tightlist
  \item
    \(1\): maximally smilar
  \item
    \(-1\): opposites
  \item
    \(0\): dissimilar
  \end{itemize}
\item
  \(\mathsf{cosineDistance}\in (0, 2)\)
\end{itemize}
\end{frame}

\begin{frame}{Cosine similarity \& distance}
\protect\hypertarget{cosine-similarity-distance-1}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/philosophy-1} \end{center}

\begin{block}{The only ``jobs'' in top-tens}
\protect\hypertarget{the-only-jobs-in-top-tens}{}
\begin{itemize}
\item
  Man: robber (.55)
\item
  Woman: policewoman (.6)
\end{itemize}
\end{block}
\end{frame}

@@

\begin{block}{No design considerations to sample size}
\protect\hypertarget{no-design-considerations-to-sample-size}{}
\begin{itemize}
\tightlist
\item
  \protect\hyperlink{ref-Ethayarajh2020measuring}{Ethayarajh}
  (\protect\hyperlink{ref-Ethayarajh2020measuring}{2020}) criticizes
  WEAT uses Bernstein bounds and argues that we would need a bias
  specific dataset of size at least 11903 to claim that the system is
  biased (three times larger than WinoBias).
\end{itemize}

@@

\begin{block}{The form of the definition is suspicious}
\protect\hypertarget{the-form-of-the-definition-is-suspicious}{}
\begin{itemize}
\tightlist
\item
  \protect\hyperlink{ref-Ethayarajh2019understanding}{Ethayarajh,
  Duvenaud, \& Hirst}
  (\protect\hyperlink{ref-Ethayarajh2019understanding}{2019}) show that
  if there are two target words only WEAT is always maximal in one
  direction.
\end{itemize}

@@

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-2}{}
\begin{block}{Our neutral words (examples, full list size = ADD)}
\protect\hypertarget{our-neutral-words-examples-full-list-size-add}{}
liquor, pow, ballpark, glitchy, billy, dallas, rip, called, outlooks,
viet, floater, rattlesnake, exports, peruvian, recursion, shortfall,
corrected, amicable, solutions, diagnostic, patently, flops, approx,
percents, lox, catapults, hamburger, engulfed, households, north,
snubbed, playtest
\end{block}

\begin{block}{Our human-related words (examples, full list size = ADD)}
\protect\hypertarget{our-human-related-words-examples-full-list-size-add}{}
switch, studio, stick, soup, sometimes, signal, prior, plant, photo,
path, park, near, menu, latter, grass, clock, wear, walk, visitor, toy,
tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character,
candidate, breakfast, supper, dinner, eat, drink, carry, run, cast, ask,
awake, ear, nose, lunch
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-3}{}
\begin{block}{Outliers and surprisingly dissimilar words}
\protect\hypertarget{outliers-and-surprisingly-dissimilar-words}{}
We study those by visualizations and uncertainty estimates.
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-4}{}
\begin{block}{Distances for ``muslim''}
\protect\hypertarget{distances-for-muslim}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-5}{}
\begin{block}{Distances for ``priest''}
\protect\hypertarget{distances-for-priest}{}
\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
\end{block}
\end{frame}

\begin{frame}{Some methodological problems}
\protect\hypertarget{some-methodological-problems-6}{}
\begin{block}{No principled interpretation}
\protect\hypertarget{no-principled-interpretation}{}
\begin{longtable}[]{@{}lllll@{}}
\toprule
Category & Biased & Hard Debiased & Soft Debiased & Diff \\
\midrule
\endhead
Religion & 0.859 & 0.934 & 0.894 & 0.075 \\
Race & 0.892 & 0.925 & 0.985 & 0.033 \\
Gender & 0.623 & 0.700 & 0.747 & 0.077 \\
\bottomrule
\end{longtable}

@@

\leavevmode\hypertarget{ref-Ethayarajh2020measuring}{}%
Ethayarajh, K. (2020). \emph{Is your classifier actually biased?
Measuring fairness under uncertainty with bernstein bounds}.

\leavevmode\hypertarget{ref-Ethayarajh2019understanding}{}%
Ethayarajh, K., Duvenaud, D., \& Hirst, G. (2019). Understanding
undesirable word embedding associations. \emph{Proceedings of the 57th
Annual Meeting of the Association for Computational Linguistics},
1696--1705.
